Differential invariant signatures for planar Lie group transformations of images

### 2020-08-13 paper/newpaper.tex
Derivation essentially there

\subsection{$SE(2)$}
An element of the special Euclidean group $SE(2)$ acts on an element $(x, y)$ of $\mathbb{R}^2$ to produce $(\bar{x}, \bar{y})$ as 
\begin{equation*}
  \begin{aligned}
    \bar{x} &= x\cos\theta  - y\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
Note that once $\theta$ has been fixed, it's always possible to choose $t_x$ and $t_y$ to ensure that $\bar{x} = \bar{y} = 0$. Prolonging the group action up to second derivatives via the chain rule gives
\begin{equation}
  \begin{bmatrix}
  \bar f_{\bar x} \\ \bar f_{\bar y} \\ \bar f_{\bar{x}\bar{x}} \\ \bar f_{\bar{x}\bar{y}} \\ \bar f_{\bar{y}\bar{y}}
  \end{bmatrix}
  \begin{bmatrix}
 c & -s & 0 & 0 & 0 \\
 s & c & 0 & 0 & 0 \\
0 & 0 & c^2 & -2cs & s^2 \\
0 & 0 & cs & c^2 - s^2 & -cs \\
0 & 0 & s^2 & 2cs & c^2 \\
  \end{bmatrix}
  \begin{bmatrix}
f_x \\ f_y \\ f_{xx} \\ f_{xy} \\ f_{yy} 
  \end{bmatrix},
\label{eqn:SE2}
 \end{equation}
where $c = \cos \theta$ and $s = \sin \theta$. We are going to choose the
moving frame to project a point in the jet space onto the submanifold
$\bar{x} = 0$, $\bar{y} = 0$, $\bar{f}_{\bar{y}} = 0$, $\bar{f}_{\bar{x}} \ge
0$. This can be thought of as rotating the image about $(\bar{x}, \bar{y})$
so that the gradient is pointing in the positive $\bar{x}$-direction, and
then translating to the origin to put it in canonical position.

This choice gives $\cos\theta = f_x / \lVert \nabla f \rVert$ and $\sin\theta = -f_y / \lVert \nabla f \rVert$, and then $t_x, t_y$ can be chosen as above. The remaining derivatives, transformed into canonical position, are therefore invariant and can be found by substituting in $c$ and $s$. Note we square $\bar{f}_{\bar{x}}$ for neatness.
\begin{align}
  I_0 &= f \\
  I_1 &= f_x^2 + f_y^2 & &(= \bar{f}_{\bar{x}}^2)\\
  I_2 &= f_x^2 f_{xx} + 2 f_x f_y f_{xy} + f_y^2 f_{yy} &&(= \bar{f}_{\bar{x}\bar{x}}) \\
  I_3 &= f_x f_y (f_{yy} - f_{xx}) + f_{xy} (f_x^2 - f_y^2) &&(= \bar{f}_{\bar{x}\bar{y}})\\
  I_4 &= f_y^2 f_{xx} - 2 f_x f_y f_{xy} + f_x^2 f_{yy} &&(= \bar{f}_{\bar{y}\bar{y}})
\end{align}

@@

\subsection{The Euclidean Group E(2)}
An element of the Euclidean group $E(2)$ maps an element $(x, y)$ of $\mathbb{R}^2$ to $(\bar{x}, \bar{y}$ as 
\begin{equation*}
  \begin{aligned}
    \bar{x} &=  x\varepsilon\cos\theta  - y\varepsilon\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
where $\varepsilon \in \{-1, 1\}$ and the remaining parameters are the same as for $SE(2)$. If $\varepsilon = 1$ it is a rigid transformation, and if $\varepsilon = -1$ it contains a reflection.

This time, the group action is not free on $J^1$ (with elements of the form $(x, y, f, f_x, f_y)$) because there are two possible transformations that map to the normalised coordinates $(0, 0, \bar{f}, \bar{f}_{\bar{x}}, 0)$ where $\bar{f}_{\bar{x}} \ge 0$, one that reflects the gradient, and one that rotates. And first derivatives give no information about $\varepsilon$.

So, we prolong to second derivatives, giving
\begin{equation}
  \begin{bmatrix}
  \bar f_{\bar x} \\ \bar f_{\bar y} \\ \bar f_{\bar{x}\bar{x}} \\ \bar f_{\bar{x}\bar{y}} \\ \bar f_{\bar{y}\bar{y}}
  \end{bmatrix}
  \begin{bmatrix}
 \varepsilon c & -\varepsilon s & 0 & 0 & 0 \\
 s & c & 0 & 0 & 0 \\
0 & 0 & c^2 & -2cs & s^2 \\
0 & 0 & \varepsilon cs & \varepsilon(c^2 - s^2) & -\varepsilon cs \\
0 & 0 & s^2 & 2cs & c^2 \\
  \end{bmatrix}
  \begin{bmatrix}
f_x \\ f_y \\ f_{xx} \\ f_{xy} \\ f_{yy} 
  \end{bmatrix}
\label{eqn:E2prolongation}
 \end{equation}

>>>>

\subsection{The Special Euclidean Group $SE(2)$}
An element of the special Euclidean group $SE(2)$ acts on an element $(x, y)$ of $\mathbb{R}^2$ to produce $(\bar{x}, \bar{y})$ as 
\begin{equation*}
  \begin{aligned}
    \bar{x} &= x\cos\theta  - y\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
We prolong the group action to form the jet space $J^2$ with coordinates $(x,
y, f, f_x, f_y, f_{xx}, f_{xy}, f_{yy})$ via the chain rule. The
transformation rule for the derivatives is given by
\begin{equation}
  \begin{bmatrix}
  \bar f_{\bar x} \\ \bar f_{\bar y} \\ \bar f_{\bar{x}\bar{x}} \\ \bar f_{\bar{x}\bar{y}} \\ \bar f_{\bar{y}\bar{y}}
  \end{bmatrix} = 
  \begin{bmatrix}
 c & -s & 0 & 0 & 0 \\
 s & c & 0 & 0 & 0 \\
0 & 0 & c^2 & -2cs & s^2 \\
0 & 0 & cs & c^2 - s^2 & -cs \\
0 & 0 & s^2 & 2cs & c^2 \\
  \end{bmatrix}
  \begin{bmatrix}
f_x \\ f_y \\ f_{xx} \\ f_{xy} \\ f_{yy} 
  \end{bmatrix},
\label{eqn:SE2}
 \end{equation}
where $c = \cos \theta$ and $s = \sin \theta$. Because the group has three
parameters, we choose the cross section of the group orbits to be $\bar{x}
= 0, \bar{y} = 0, \bar{f}_{\bar{y}} = 0$. We also require that $\bar{f}_{x}
> 0$ to uniquely define the moving frame. Note that the group action is
locally free (away from critical points) when prolonged to $J^1$, however
we prolonged to $J^2$ in order to compute invariants. points. The moving
frame is then given by the prolonged Euclidean transformation that maps an
element of  $J^2$ to this cross section. It can be thought of as rotating
the image about that point so that its gradient is pointing in the positive
$x$ direction, and then translating the image so the point is mapped to the
origin.

This choice of moving frame gives $\cos\theta = f_x/\lVert\nabla f\rVert$
and $\sin\theta = -f_y/\lVert \nabla f \rVert$. The parameters $t_x, t_y$
are then $t_x = -x\cos\theta + y\sin\theta$ and $t_y = -x\sin\theta -
y\cos\theta$, however these have no effect on the derivatives, and will
henceforth be ignored.

The remaining elements not fixed by the cross section, $\bar{f}$,
$\bar{f}_{\bar{x}}$, $\bar{f}_{\bar{x}\bar{x}}$,
$\bar{f}_{\bar{x}\bar{y}}$, and $\bar{f}_{\bar{y}\bar{y}}$ are therefore
all invariant. This gives the following set of invariants:
\begin{equation}\label{eq:se2invariants}
\begin{aligned}
  I_0 &= f \\
  I_1 &= f_x^2 + f_y^2 & &(= \bar{f}_{\bar{x}}^2)\\
  I_2 &= f_x^2 f_{xx} + 2 f_x f_y f_{xy} + f_y^2 f_{yy} &&(= I_1\bar{f}_{\bar{x}\bar{x}}) \\
  I_3 &= f_x f_y (f_{yy} - f_{xx}) + f_{xy} (f_x^2 - f_y^2) &&(= I_1\bar{f}_{\bar{x}\bar{y}})\\
  I_4 &= f_y^2 f_{xx} - 2 f_x f_y f_{xy} + f_x^2 f_{yy} &&(= I_1\bar{f}_{\bar{y}\bar{y}})
\end{aligned}
\end{equation}
Note that we multiplied each of the second derivative invariants by $I_1$
so that they would decay smoothly to zero at critical points.

@@

\subsection{The Euclidean Group E(2)}
An element of the Euclidean group $E(2)$ maps an element $(x, y)$ of
$\mathbb{R}^2$ to $(\bar{x}, \bar{y})$ as 
\begin{equation*}
  \begin{aligned}
    \bar{x} &=  x\varepsilon\cos\theta  - y\varepsilon\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
where $\varepsilon \in \{-1, 1\}$ and the remaining parameters are the same
as for $SE(2)$. If $\varepsilon = 1$ it is a rigid transformation, and if
$\varepsilon = -1$ it contains a reflection.

This time, the group action is not free on $J^1$ (with elements of the form
$(x, y, f, f_x, f_y)$) away from critical points because there are two
possible transformations that map to the cross section $(0, 0, \bar{f},
\bar{f}_{\bar{x}}, 0)$ where $\bar{f}_{\bar{x}} > 0$, one that reflects the
gradient into the positive $x$ direction, and one that rotates.

We again prolong the group action to $J^2$, but this time we expect to need
to use a second derivative term to determine $\varepsilon$ for the moving
frame. The derivative transformations are given by
\begin{equation}
  \begin{bmatrix}
  \bar f_{\bar x} \\ \bar f_{\bar y} \\ \bar f_{\bar{x}\bar{x}} \\ \bar f_{\bar{x}\bar{y}} \\ \bar f_{\bar{y}\bar{y}}
  \end{bmatrix} = 
  \begin{bmatrix}
 \varepsilon c & -\varepsilon s & 0 & 0 & 0 \\
 s & c & 0 & 0 & 0 \\
0 & 0 & c^2 & -2cs & s^2 \\
0 & 0 & \varepsilon cs & \varepsilon(c^2 - s^2) & -\varepsilon cs \\
0 & 0 & s^2 & 2cs & c^2 \\
  \end{bmatrix}
  \begin{bmatrix}
f_x \\ f_y \\ f_{xx} \\ f_{xy} \\ f_{yy} 
  \end{bmatrix},
\label{eqn:E2prolongation}
\end{equation}
where again $c = \cos\theta$ and $s = \sin\theta$.

We begin by choosing the same cross section as for $SE(2)$, namely $\bar{x} = 0,
\bar{y} = 0, \bar{f}_{\bar{y}} = 0, \bar{f}_{\bar{x}} > 0$. This gives
$\cos\theta = \varepsilon f_x / \lVert \nabla f \rVert$ and $\sin\theta =
-\varepsilon f_y / \lVert \nabla f \rVert$. Under this transformation,
regardless of whether $\varepsilon = 1$ or $\varepsilon = -1$, the
derivative $\bar{f}_{\bar{x}} = \lVert \nabla f \rVert$. We need a second
derivative term to resolve the choice of $\varepsilon$ for the moving
frame. The only second derivative term in which $\varepsilon$ appears is
the $\bar{f}_{\bar{x}\bar{y}}$ term. Substituting in for $\cos\theta,
\sin\theta$, we see that
\begin{equation*}
    \bar{f}_{\bar{x}\bar{y}} = \frac{\varepsilon(f_xf_y(f_{yy} - f_{xx}) + (f_x^2 -
    f_y^2)f_{xy})}{f_x^2 + f_y^2}
\end{equation*}
This suggests choosing $\varepsilon$ to make $\bar{f}_{\bar{x}\bar{y}} \ge
0$. We can then list the invariants:
\begin{equation}\label{eq:e2invariants}
\begin{aligned}
  I_0 &= f \\
  I_1 &= f_x^2 + f_y^2 & &(= \bar{f}_{\bar{x}}^2)\\
  I_2 &= f_x^2 f_{xx} + 2 f_x f_y f_{xy} + f_y^2 f_{yy} &&(= I_1\bar{f}_{\bar{x}\bar{x}}) \\
  I_3 &= \lvert f_x f_y (f_{yy} - f_{xx}) + f_{xy} (f_x^2 - f_y^2)\rvert
  &&(= I_1\lvert \bar{f}_{\bar{x}\bar{y}}\rvert)\\
  I_4 &= f_y^2 f_{xx} - 2 f_x f_y f_{xy} + f_x^2 f_{yy} &&(= I_1\bar{f}_{\bar{y}\bar{y}})
\end{aligned}
\end{equation}
Notice that most, but not all of the $SE(2)$ invariants
\eqref{eq:se2invariants} are also $E(2)$ invariants. We again multiply
through the invariants by $I_1$ so that they decay smoothly at critical
points.
